### 2025-02-12 paper/sections/intro.qmd

Today's prominence of artificial intelligence (AI) has largely been driven by advances in **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern AIs are tasked with learning these representations from scratch, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep]. Modern advances in computing have made it possible to provide such AIs with ever greater degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models. Unfortunately, in doing so they also learn increasingly complex and highly sensitive representations that we can no longer easily interpret.

@@

Part of the problem is that high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case]. Based purely on predictive performance, these solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns. This poses two related challenges: firstly, it makes these models inherently opaque, since humans cannot simply interpret what type of explanation the complex learned representations correspond to; secondly, even if we could resolve the first challenge, it is not obvious how to mitigate models from learning representations that correspond to meaningless and undesirable explanations.

@@

To our surprise, the second challenge has not yet attracted any consolidated research effort. Specifically, there has been no concerted effort towards improving model **explainability**, which we define here as the degree to which learned representations correspond to explanations that are interpretable and deemed desirable by humans. Instead, the choice has typically been to improve the capacity of XAI tools to identify the subset explanations that are both desirable and valid for any given model, independent of whether the learned representations are also compatible with undesirable explanations [@altmeyer2024faithful]. Fortunately, recent findings indicate that explainability can arise as byproduct of regularization techniques aimed at other objectives such as robustness, generalization and generative capacity [@schut2021generating, @augustin2020adversarial, @altmeyer2024faithful].

Building on these findings, we introduce **counterfactual training**: a novel regularization technique geared explicitly towards aligning model representations with desirable explanations. Our contributions are as follows:

>>>>

Today's prominence of artificial intelligence (AI) has largely been driven by advances in **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern machine learning (ML) models are tasked with learning these representations from scratch, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep]. Modern advances in computing have made it possible to provide such models with ever greater degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models. Unfortunately, in doing so they also learn increasingly complex and highly sensitive representations that we can no longer easily interpret.

@@

Part of the problem is that high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case]^[For clarity: we follow standard ML convention in using "degrees of freedom" to refer to the number of parameters estimated from data.]. Based purely on predictive performance, these solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns. This poses two related challenges: firstly, it makes these models inherently opaque, since humans cannot simply interpret what type of explanation the complex learned representations correspond to; secondly, even if we could resolve the first challenge, it is not obvious how to mitigate models from learning representations that correspond to meaningless and implausible explanations.

@@

To our surprise, the second challenge has not yet attracted any consolidated research effort. Specifically, there has been no concerted effort towards improving model **explainability**, which we define here as the degree to which learned representations correspond to explanations that are interpretable and deemed **plausible** by humans (see @def-explainability). Instead, the choice has typically been to improve the capacity of XAI tools to identify the subset explanations that are both plausible and valid for any given model, independent of whether the learned representations are also compatible with implausible explanations [@altmeyer2024faithful]. Fortunately, recent findings indicate that explainability can arise as byproduct of regularization techniques aimed at other objectives such as robustness, generalization and generative capacity [@schut2021generating, @augustin2020adversarial, @altmeyer2024faithful].

Building on these findings, we introduce **counterfactual training**: a novel regularization technique geared explicitly towards aligning model representations with plausible explanations. Our contributions are as follows:

### 2025-02-12 paper/sections/method.qmd

From a Bayesian perspective, setting $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x})$ with respect to feature $d$. Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments. Under certain conditions, this results holds theoretically^[For the proof, see the supplementary appendix.]:

@@

Let $\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}$ denote a linear classifier with $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$. If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty [@eq-div] will result in lower classifier sensitivity to that feature, provided that at least one other feature is mutable.

>>>>

From a Bayesian perspective, setting $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x})$ with respect to feature $d$. Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments. Under certain conditions, this results holds theoretically[For the proof, see the supplementary appendix.]:

@@

Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ (i.e. *multinomial logistic regression*) where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$. If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty (@eq-div) will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is mutable.

### 2025-02-12 paper/sections/other/constraints.qmd

By @eq-loglike-decomp we have the $\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0$ and $\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0$, where $\theta_{k,d}=\Theta[k,d]$ denotes the coefficient on feature $d$ for class $k$.

@@

Since $\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}$ we are more likely to have $x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}} < 0$ than vice versa at initialization. Similarly, we are more likely to have $x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}} > 0$ by $\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}$.

This implies that if we do not protect feature $d_{\text{immtbl}}$, the contrastive divergence penalty will push down on $\theta_{y^-,d_{\text{immtbl}}}$ thereby exacerbating the existing effect $\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0$. In words, not protecting the immutable feature would have the undesirable effect of making the classifier more sensitive to this feature, in that it would be more likely to predict class $y^-$ as opposed to $y^+$ for lower values of $d_{\text{immtbl}}$.

By the same rationale, the contrastive divergence penalty can generally be expected to push up on $\theta_{y^-,d_{\text{mtbl}}}$ exacerbating $\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0$. In words, this has the effect of making the classifier more sensitive to the mutable feature, in that it would be more likely to predict class $y^-$ as opposed to $y^+$ for higher values of $d_{\text{mtbl}}$.

>>>>

By the properties of our classifier (*multinomial logistic regression*), we have:

$$
\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \left( \theta_{k,d} - \theta_{m,d} \right)x_d + \text{const}
$$ {#eq-multi}

where $\theta_{k,d}=\Theta[k,d]$ denotes the coefficient on feature $d$ for class $k$.

Based on @eq-loglike-decomp and @eq-multi we can identify that $(\mu_{k,d}-\mu_{m,d}) \propto (\theta_{k,d} - \theta_{m,d})$ under the assumptions we made above. Hence, we have that $(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0$ and $(\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0$

@@

Since $(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}})$ we are more likely to have $(x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}}) < 0$ than vice versa at initialization. Similarly, we are more likely to have $(x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}}) > 0$ since $(\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}})$.

This implies that if we do not protect feature $d_{\text{immtbl}}$, the contrastive divergence penalty will push down on $\theta_{y^-,d_{\text{immtbl}}}$ thereby exacerbating the existing effect $(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0$. In words, not protecting the immutable feature would have the undesirable effect of making the classifier more sensitive to this feature, in that it would be more likely to predict class $y^-$ as opposed to $y^+$ for lower values of $d_{\text{immtbl}}$.

By the same rationale, the contrastive divergence penalty can generally be expected to push up on $\theta_{y^-,d_{\text{mtbl}}}$ exacerbating $(\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0$. In words, this has the effect of making the classifier more sensitive to the mutable feature, in that it would be more likely to predict class $y^-$ as opposed to $y^+$ for higher values of $d_{\text{mtbl}}$.

### 2025-02-12 paper/sections/related.qmd

To the best of our knowledge, our proposed framework for counterfactual training represents the first attempt to use counterfactual explanations during training to improve model explainability. In high-level terms, we define model explainability as the extent to which valid explanations derived for an opaque model are also deemed desirable with respect to the underlying data and stakeholder requirements. To make this more concrete, we follow @augustin2020adversarial in tieing the concept of explainability to the quality of counterfactual explanations that we can generate for a given model. The authors show that counterfactual explanations---understood here as minimal input perturbations that yield some desired model prediction---are generally more meaningful if the underlying model is more robust to adversarial examples. We can make intuitive sense of this finding when looking at adversarial training (AT) through the lens of representation learning with high degrees of freedom: by inducing models to "unlearn" representations that are susceptible to worst-case counterfactuals (i.e. adversarial examples), AT effectively removes some undesirable explanations from the solution space.

@@

Improving the adversarial robustness of models is not the only path towards aligning representations with desirable explanations. In a work closely related to this one, @altmeyer2024faithful show that explainability can be improved through model averaging and refined model objectives. The authors propose a way to generate counterfactuals that are maximally **faithful** to the model in that they are consistent with what the model has learned about the underlying data. Formally, they rely on tools from energy-based modelling to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model. Their proposed counterfactual explainer, *ECCCo*, yields desirable (or **plausible**) explanations if and only if the underlying model has learned representations that align with them. They find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your] tend to do well in this regard.

Once again it helps to look at these findings through the lens of representation learning with high degrees of freedom. Deep ensembles are approximate Bayesian model averages, which are most called for when models are underspecified by the available data [@wilson2020case]. Averaging across solutions mitigates the aforementioned risk of relying on a single locally optimal representations that corresponds to semantically meaningless explanations for the data. Previous work by @schut2021generating similarly found that generating desirable ("interpretable") counterfactual explanations is almost trivial for deep ensembles that have also undergone adversarial training. The case for JEMs is even clearer: they involve a hybrid objective that induces both high predictive performance and generative capacity [@grathwohl2020your]. This is closely related to the idea of aligning models with desirable explanations and has inspired our proposed counterfactual training objective, as we explain in @sec-method.

>>>>

To the best of our knowledge, our proposed framework for counterfactual training represents the first attempt to use counterfactual explanations during training to improve model explainability. In high-level terms, we define model explainability as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and stakeholder requirements. To make this more concrete, we follow @augustin2020adversarial in tieing the concept of explainability to the quality of counterfactual explanations that we can generate for a given model. The authors show that counterfactual explanations---understood here as minimal input perturbations that yield some desired model prediction---are generally more meaningful if the underlying model is more robust to adversarial examples. We can make intuitive sense of this finding when looking at adversarial training (AT) through the lens of representation learning with high degrees of freedom: by inducing models to "unlearn" representations that are susceptible to worst-case counterfactuals (i.e. adversarial examples), AT effectively removes some implausible explanations from the solution space.

@@

Improving the adversarial robustness of models is not the only path towards aligning representations with plausible explanations. In a work closely related to this one, @altmeyer2024faithful show that explainability can be improved through model averaging and refined model objectives. The authors propose a way to generate counterfactuals that are maximally **faithful** to the model in that they are consistent with what the model has learned about the underlying data. Formally, they rely on tools from energy-based modelling to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model. Their proposed counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them. They find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your] tend to do well in this regard.

Once again it helps to look at these findings through the lens of representation learning with high degrees of freedom. Deep ensembles are approximate Bayesian model averages, which are most called for when models are underspecified by the available data [@wilson2020case]. Averaging across solutions mitigates the aforementioned risk of relying on a single locally optimal representations that corresponds to semantically meaningless explanations for the data. Previous work by @schut2021generating similarly found that generating plausible ("interpretable") counterfactual explanations is almost trivial for deep ensembles that have also undergone adversarial training. The case for JEMs is even clearer: they involve a hybrid objective that induces both high predictive performance and generative capacity [@grathwohl2020your]. This is closely related to the idea of aligning models with plausible explanations and has inspired our proposed counterfactual training objective, as we explain in @sec-method.
